Galactic Chemical Evolution Models Favor an Extended Type Ia Supernova Delay-Time Distribution

### 2024-03-18 src/tex/ms.tex
[showyourwork] overleaf sync

\shorttitle{SN Ia DTD in GCE Simulations}
\shortauthors{Dubay, Johnson, \& Johnson}
% \linespread{1.8}

@@

% Abstract from AAS 241
\begin{abstract}
    Type Ia supernovae (SNe Ia) produce most of the Fe-peak elements in the Universe and therefore are a crucial ingredient in galactic chemical evolution models. SNe Ia do not explode immediately after star formation and the delay-time distribution (DTD) has not been definitively determined by supernova surveys or theoretical models. Because the DTD also affects the relationship among age, [Fe/H], and \aFe in chemical evolution models, comparison with observations of stars in the Milky Way is an important consistency check for any proposed DTD. We implement several popular forms of the DTD in combination with multiple star formation histories for the Milky Way in one-zone and multi-zone chemical evolution models using the Versatile Integrator for Chemical Evolution (\vice). \vice includes a prescription for radial stellar migration in multi-zone models. We compare our predicted interstellar medium abundance tracks, stellar abundance distributions, and stellar age distributions to the final data release of the Apache Point Observatory Galactic Evolution Experiment (APOGEE). We find that the SN Ia DTD has the largest effect on the \aFe distribution: a DTD with more prompt SNe Ia produces a stellar abundance distribution that is skewed toward a lower \aFe ratio, whereas a DTD with more delayed SNe Ia produces a distribution that is skewed toward higher \aFe. While the DTD alone cannot explain the observed bimodality in the \aFe distribution, in combination with an appropriate star formation history it affects the goodness of fit between the simulated and observed high-$\alpha$ sequence. Our model results favor a DTD with longer delay times and fewer prompt SNe Ia than the fiducial $t^{-1}$ power law.
\end{abstract}

@@

% [Hook and intro paragraph]
Galactic chemical evolution (GCE) studies seek to explain the observed distribution of metals throughout the Milky Way Galaxy. \citet{Tinsley1979-StellarLifetimes} made a compelling case that the non-Solar \aFe\footnote{
    In standard bracket notation, $[X/Y]\equiv \log_{10}(X/Y) - \log_{10}(X/Y)_{\odot}$. In this work we will use \aFe and [O/Fe] interchangeably, although observational studies will often use a combination of $\alpha$-elements to calculate \aFe.
} ratios seen by, e.g., \citet{Wallerstein1962-GDwarfAbundances} were caused by different stellar lifetimes for the contributors of the Fe-peak elements than the $\alpha$-elements. 
Type Ia supernovae (SNe Ia), the thermonuclear explosions of carbon-oxygen white dwarfs (WDs), are responsible for a majority of the Fe produced in the Galaxy \citep{Matteucci1986-SupernovaEnrichment}; meanwhile, core collapse supernovae (CCSNe), the explosions of massive stars, produce the $\alpha$-elements (e.g., O and Mg) in addition to a smaller fraction of Fe.
SNe Ia are delayed by $\sim10^8 - 10^{10}$ years after star formation events, as evidenced by observations at large distances from star forming regions and in elliptical galaxies \citep[e.g.,][]{Maza1976-SNStatistics}.
This delayed enrichment leads to a decrease in \aFe with increasing [Fe/H] \citep{Matteucci1986-SupernovaEnrichment}.
% The decrease in \aFe with [Fe/H] has long been understood to be a consequence of the delayed nature of SNe Ia, which do not contribute substantially to chemical enrichment for $\sim 10^8$ years after the commencement of star formation \citep{Tinsley1979-StellarLifetimes,Matteucci1986-SupernovaEnrichment}. 
% As SNe Ia are a product of intermediate-mass stars which undergo binary evolution, they do not contribute substantially to chemical enrichment for $\sim 10^8$ years after the commencement of star formation \citep{Matteucci1986-SupernovaEnrichment}.
% The timescale for Fe enrichment in the Galaxy, a key component of chemical evolution models, is directly linked to the progenitor lifetimes of SNe Ia. 
Therefore, the relative abundances of the $\alpha$-elements and Fe as a function of stellar age trace the balance of SN rates over time.

% [What is the delay time distribution]
The rate of SNe Ia which explode as a function of time after the formation of a hypothetical unit-mass single stellar population is known as the delay-time distribution \citep[DTD; for a review, see Section 3.5 of][]{Maoz2014-Review}.
When the DTD is convolved with the Galactic star formation rate (SFR) as a function of time, it yields the SN Ia rate as a function of time. 
The quantitative details of the relationship between \aFe and [Fe/H] are set by the DTD, and as such it is a key parameter in GCE models.
The DTD remains poorly constrained, however, because it reflects the detailed evolution of the SNe Ia progenitor systems, so different models for the progenitors of SNe Ia will naturally predict different forms for the DTD.

% [Introduction to SNe Ia]
The mechanism(s) for producing SNe Ia are not fully understood \citep[for reviews, see][]{Maoz2014-Review,Livio2018-ProgenitorReview,Ruiter2020-ProgenitorReview}. Two general production channels have been proposed. In the single-degenerate (SD) case, the WD accretes mass from a close non-degenerate companion until it surpasses $\sim1.4$ M$_\odot$ and explodes \citep{Whelan1973-SDModel,Nomoto1982-SDModel,Yoon2003-SDModel}. In the double-degenerate (DD) case, two WDs merge after a gravitational-wave inspiral \citep{Iben1984-IaBinary,Webbink1984-DDModel,Pakmor2012-WDMerger} or head-on collision \citep{Benz1989-CollisionalDD,Thompson2011-CollisionalDD}. Searches for signs of interaction between the SN ejecta and a non-degenerate companion \citep[e.g.,][]{Panagia2006-RadioEmission,Chomiuk2016-RadioEmission,Fausnaugh2019-EarlyIaLightCurves,Tucker2020-SNeIaSpectra,Dubay2022-SNeIaCSM} or for a surviving companion \citep[e.g.,][]{Schaefer2012-ExCompanionSNR,Do2021-Progenitor1972E,Tucker2023-SN2011fe} have placed tight constraints on the SD channel, heavily disfavoring it as the main pathway for producing ``normal'' SNe Ia. The DD channel, now the preferred model, faces issues of its own with matching observed SN Ia rates because not all WD mergers may lead to a thermonuclear explosion \citep[e.g.,][]{NomotoIben1985-DDMergers,SaioNomoto1998-DDMergers,Shen2012-DDMergers}, and the progenitor systems are difficult to detect even within our own Galaxy \citep{RebassaMansergas2019-WhereAreDDProgenitors}.

% [Theoretical DTDs]
Uncertainties in the progenitor channels of SNe Ia translate into uncertainties in the shape of the DTD. 
% By formulating analytical DTDs for several progenitor models, \citet{Greggio2005-AnalyticalRates} found that the SD channel tends to produce a steeper DTD function than the DD channel, and the median delay time (by which 50\% of SNe Ia have exploded) can vary by a factor of 10 depending on the chosen progenitor scenario and its particular parameters.
For the DD channel, assumptions about the distribution of post-common envelope separations and the rate of gravitational wave inspiral suggest a broad $\sim t^{-1}$ DTD at long delay times ($\gtrsim 1$ Gyr), but at short delays ($\lesssim 1$ Gyr) the rate is limited by the need to produce two WDs \citep[see Section 3.5 from][]{Maoz2014-Review}.
Higher-order progenitor systems could also produce a $t^{-1}$ DTD \citep{Fang2018-QuadrupleSystems,Rajamuthukumar2023-TripleEvolution}.
The DTD which would result from the SD channel depends greatly on the assumptions of binary population synthesis, but in general is expected to cover a narrower range of delay times and may feature a steep exponential cutoff at the long end \citep[e.g.,][]{Greggio2005-AnalyticalRates}.
% More recently, \citet{Rajamuthukumar2023-TripleEvolution} computed a DTD from the evolution of triple system progenitors. [Expand, connect binary evolution to delay times, look for other theory papers]

% [Observational DTDs]
Surveys of SNe Ia have provided the primary observational constraints on the DTD.  
This can be done by comparing the observed rate of SNe Ia to their host galaxy parameters \citep[e.g.,][]{Mannucci2005-SNRate,Heringer2019-FieldGalaxyDTD} or inferred star formation histories \citep[SFHs; e.g.,][]{Maoz2012-SloanIIDTD}, measuring SN Ia rates in galaxy clusters \citep[e.g.,][]{Maoz2010-ClusterDTD}, or comparing the volumetric SN Ia rate to the cosmic SFH as a whole \citep[e.g.,][]{Graur2014-VolumetricSNIaRates,Strolger2020-ExponentialDTD}. 
Early studies, which had limited sample sizes, produced unimodal \citep{Strolger2004-SNIaProgenitors} or bimodal \citep{Mannucci2006-TwoPopulations} DTDs where the majority of SNe Ia explode within a relatively narrow range of delay times.
More recent studies have recovered broader DTD functions, with many converging on a declining power-law of $\sim t^{-1}$ \citep[e.g.,][]{Maoz2017-CosmicDTD,Castrillo2021-DTD,Wiseman2021-DESRates}, though there is some evidence for a steeper slope in galaxy clusters \citep{Maoz2017-CosmicDTD,FriedmannMaoz2018-ClusterDTD}.
It is especially difficult to constrain the DTD for short delay times \citep{MaozMannucci2012-SNeIaReview} because of the need for SN Ia rates at long lookback times and uncertainties in the age estimates of stellar populations.

@@

The uncertainties in the SN Ia DTD propagate into GCE models. In principle, then, the observed chemical abundance patterns should contain information about the DTD, and therefore the progenitors, of SNe Ia.
A few studies have compared different DTDs in one-zone chemical evolution models, but their comparisons to abundance data are limited to the local Solar neighborhood \citep[e.g.,][]{Andrews2017-ChemicalEvolution,Palicio2023-AnalyticDTD}. 
[Talk about \citep{Matteucci2009-DTDModels} multi-zone models].
\citet{Poulhazan2018-PrecisionPollution} compared six DTD models in a smoothed-particle hydrodynamics simulation in a cosmological context. They found that models with a prompt SN Ia component produced the highest abundance ratios and the narrowest \aFe distributions at late times, although their simulation was not tuned to reproduce the parameters of the Milky Way and they did not compare to observed chemical abundances. 
The current era of large spectroscopic surveys such as the Apache Point Observatory Galactic Evolution Experiment \citep[APOGEE;][]{Majewski2017-APOGEE} and the forthcoming Milky Way Mapper [cite] has made abundances across the Milky Way disk available for comparison to more sophisticated GCE models.

% Notes on GCE studies
% \citet{Andrews2017-ChemicalEvolution} predict abundance tracks and distributions using one-zone models for four different DTD models; they find that most are consistent with their abundance data.
% \citet{Palicio2023-AnalyticDTD} derive analytic solutions to one-zone chemical evolution equations for seven DTD models, but their comparisons to data are limited to the Solar neighborhood. 
% \citet{Poulhazan2018-PrecisionPollution} investigate six different DTD models in a smoothed-particle hydrodynamics simulation, but they only present global results for the entire galaxy, they don't compare abundances throughout the disk, and they don't compare to observations.

This paper presents the first comprehensive look at the DTD in a multi-zone GCE model that can qualitatively reproduce the observed abundance structure of the Milky Way disk. A multi-zone approach allows for a radially-dependent parameterization of the SFH, outflows, stellar migration, and abundance gradient which can better match observations across the Galactic disk. We evaluate a selection of DTD models from the literature with multiple SFHs and a prescription for radial stellar migration in the Versatile Integrator for Chemical Evolution \citep[\vice;][]{JohnsonWeinberg2020-Starbursts}. In Section \ref{sec:methods}, we present our models for the DTD and SFH, and we describe our chemical evolution models. In Section \ref{sec:onezone-results}, we present the results of our one-zone chemical evolution models. In Section \ref{sec:multizone-results}, we present the results of our multi-zone models and compare to observations. In Section \ref{sec:discussion}, we discuss the implications for the DTD and future surveys. In Section \ref{sec:conclusions}, we summarize our conclusions.

@@

[Quick note about mass loading?]

@@

For simplicity and easier comparison to the results of \citet{Johnson2021-Migration} (hereafter \citetalias{Johnson2021-Migration}), we focus our analysis on O and Fe, representing the $\alpha$ and Fe-peak elements, respectively. Both elements are produced by CCSNe. \vice adopts the instantaneous recycling approximation for CCSNe, so the equation which governs CCSN enrichment as a function of star formation for some element $x$ is simply
\begin{equation}
    \dot M_x^{\rm CC} = y_x^{\rm CC} \dot M_*
    \label{eq:ccsn-enrichment}
\end{equation}
where $y_x^{\rm CC}$ is the CCSN yield of element $x$ per unit mass of star formation, and $\dot M_*$ is the SFR. 
Following \citetalias{Johnson2021-Migration}, who in turn adopt their CCSN yields from \citet{ChieffiLimongi2004-CCSNYields} and \citet{LimongiChieffi2006-CCSNYields}, we adopt $y_{\rm O}^{\rm CC}=0.015$ and $y_{\rm Fe}^{\rm CC}=0.0012$. The primary effect of these yields is to set the low-[Fe/H] ``plateau'' in [O/Fe] which represents pure CCSN enrichment. The chosen yields for this paper produce a plateau at ${\rm [O/Fe]}=0.45$; see \citet{Weinberg2023-CCSNYield} for more discussion on the effect of the CCSN yields on chemical evolution.

Following the formalism of \citet{Weinberg2017-ChemicalEquilibrium}, the rate of Fe contribution to the ISM from SNe Ia is 
\begin{equation}
    \dot M_{\rm Fe}^{\rm Ia} = y_{\rm Fe}^{\rm Ia} \langle \dot M_*\rangle_{\rm Ia} \\
\end{equation}
where $\langle \dot M_*\rangle_{\rm Ia}$ is the time-averaged SFR weighted by the DTD and $y_{\rm Fe}^{\rm Ia}$ is the Fe yield of SNe Ia. \citet{Weinberg2017-ChemicalEquilibrium} show in their Appendix A that
\begin{equation}
    \langle \dot M_*\rangle_{\rm Ia} \equiv \frac{\int_0^t \dot M_*(t')R_{\rm Ia}(t-t')dt'}{\int_{t_D}^{t_{\rm max}} R_{\rm Ia}(t')dt'}
    \label{eq:weighted-sfr}
\end{equation}
where %$\dot M_*(t)$ is the SFR at time $t$, 
%$R_{\rm Ia}(t) dt$ is the number of SNe Ia which explode in the time interval $(t, t+dt)$ per unit mass of stars formed at $t=0$, 
$R_{\rm Ia}$ is the DTD in units of ${\rm M}_\odot^{-1}\,{\rm yr}^{-1}$,
$t_D$ is the minimum SN Ia delay time, and $t_{\rm max}$ is the maximum simulation time. The denominator of Equation \ref{eq:weighted-sfr} is therefore the total number of SNe Ia per M$_{\odot}$ of stars formed.

The yield $y_{\rm Fe}^{\rm Ia}$ measures the mass of Fe produced by SNe Ia per unit mass of star formation. It can be expressed in terms of the integral of the DTD:
\begin{equation}
    y_{\rm Fe}^{\rm Ia} = m_{\rm Fe}^{\rm Ia} \int_{t_D}^{t_{\rm max}} R_{\rm Ia}(t')dt' = m_{\rm Fe}^{\rm Ia} \frac{N_{\rm Ia}}{M_*},
    \label{eq:dtd-integral}
\end{equation}
where $m_{\rm Fe}^{\rm Ia}$ is the average mass of Fe produced by a single SN Ia, and $N_{\rm Ia}/M_*=2.2\times10^{-3}\,{\rm M}_\odot^{-1}$ is the average number of SNe Ia per mass of stars formed \citep{MaozMannucci2012-SNeIaReview}. 
Adjusting the value of $y_{\rm Fe}^{\rm Ia}$ primarily affects the end point of chemical evolution tracks in [O/Fe]--[Fe/H] space. 
Following \citetalias{Johnson2021-Migration}, we adopt $y_{\rm Fe}^{\rm Ia}=0.00214$. This yield was originally adapted from the W70 model of \citet{Iwamoto1999-SNIaYields}, but was slightly modified so that the inside-out SFH model produces stars with ${\rm [O/Fe]}\approx 0.0$ by the end of the simulation.
\citet{Palla2021-SNIaYield} studies the effect of different SN Ia yields on GCE models in detail.

\subsection{DTD Models}
\label{sec:dtd-models}

We explore five different functional forms for the SN Ia DTD: a power-law, a broken power-law with an initial flat plateau, an exponential, a two-population model, and a model based on triple-system dynamics, with a one or two variants of each. Figure \ref{fig:dtds} presents these DTDs as a function of time after star formation, and Table \ref{tab:dtds} summarizes the DTD models and their parameters. We use simple forms rather than simulated physical or analytical models of SNe Ia for the sake of decreased computational time. Additionally, physically-motivated models of the DTD must contend with many unknown or poorly-constrained parameters, so our simple forms allow us to reduce the number of free parameters. In Appendix \ref{app:analytical-dtds}, we show that a few of our simple forms adequately approximate the more complete analytical models of \citet{Greggio2005-AnalyticalRates}.

In this subsection we present unnormalized functional forms $f_{\rm Ia}$ for each DTD model. The relationship between $f_{\rm Ia}$, which has units of ${\rm Gyr}^{-1}$, and $R_{\rm Ia}$, which has units of ${\rm M}_\odot^{-1}\,{\rm Gyr}^{-1}$, is: 
\begin{equation}
    R_{\rm Ia}(t) = 
    \begin{cases}
        \frac{N_{\rm Ia}}{M_*}
        \frac{f_{\rm Ia}(t)}{\int_{t_D}^{t_{\rm max}} f_{\rm Ia}(t') dt'}, & t \ge t_D \\
        0 & t < t_D.
    \end{cases}
    \label{eq:dtd-function}
\end{equation}
\vice automatically normalizes the provided function for the DTD according to this equation.

>>>>

\shorttitle{SN Ia DTD in GCE Models}
\shortauthors{Dubay, Johnson, \& Johnson}
% \linespread{1.8}

@@

% Abstract from AAS 241
\begin{abstract}
    Type Ia supernovae (SNe Ia) produce most of the Fe-peak elements in the Universe and therefore are a crucial ingredient in galactic chemical evolution models. SNe Ia do not explode immediately after star formation and the delay-time distribution (DTD) has not been definitively determined by supernova surveys or theoretical models. Because the DTD also affects the relationship among age, [Fe/H], and \aFe in chemical evolution models, comparison with observations of stars in the Milky Way is an important consistency check for any proposed DTD. We implement several popular forms of the DTD in combination with multiple star formation histories for the Milky Way in %one-zone and 
    multi-zone chemical evolution models including a prescription for radial stellar migration.
    % using the Versatile Integrator for Chemical Evolution (\vice). \vice includes a prescription for radial stellar migration in multi-zone models. 
    We compare our predicted interstellar medium abundance tracks, stellar abundance distributions, and stellar age distributions to the final data release of the Apache Point Observatory Galactic Evolution Experiment (APOGEE). We find that the SN Ia DTD has the largest effect on the \aFe distribution: a DTD with more prompt SNe Ia produces a stellar abundance distribution that is skewed toward a lower \aFe ratio, whereas a DTD with more delayed SNe Ia produces a distribution that is skewed toward higher \aFe. While the DTD alone cannot explain the observed bimodality in the \aFe distribution, in combination with an appropriate star formation history it affects the goodness of fit between the simulated and observed high-$\alpha$ sequence. Our model results favor a DTD with longer delay times and fewer prompt SNe Ia than the fiducial $t^{-1}$ power law.
\end{abstract}

@@

% [Hook and intro paragraph]
Galactic chemical evolution (GCE) studies seek to explain the observed distribution of metals throughout the Milky Way Galaxy. \citet{Tinsley1979-StellarLifetimes} made a compelling case that the non-solar \aFe\footnote{
    In standard bracket notation, $[X/Y]\equiv \log_{10}(X/Y) - \log_{10}(X/Y)_{\odot}$. In this work we will use \aFe and [O/Fe] interchangeably, although observational studies will often use a combination of $\alpha$-elements to calculate \aFe.
} ratios seen by, e.g., \citet{Wallerstein1962-GDwarfAbundances} were caused by different stellar lifetimes for the contributors of the Fe-peak elements than the $\alpha$-elements. 
Type Ia supernovae (SNe Ia), the thermonuclear explosions of carbon-oxygen white dwarfs (WDs), are responsible for a majority of the Fe produced in the Galaxy \citep{Matteucci1986-SupernovaEnrichment}; meanwhile, core collapse supernovae (CCSNe), the explosions of massive stars, produce the $\alpha$-elements (e.g., O and Mg) in addition to a smaller fraction of Fe.
SNe Ia are delayed by $\sim0.1-10$ Gyr after star formation events, as evidenced by observations at large distances from star forming regions and in elliptical galaxies \citep[e.g.,][]{Maza1976-SNStatistics}.
This delayed enrichment leads to a decrease in \aFe with increasing [Fe/H] \citep{Matteucci1986-SupernovaEnrichment}.
% The decrease in \aFe with [Fe/H] has long been understood to be a consequence of the delayed nature of SNe Ia, which do not contribute substantially to chemical enrichment for $\sim 10^8$ years after the commencement of star formation \citep{Tinsley1979-StellarLifetimes,Matteucci1986-SupernovaEnrichment}. 
% As SNe Ia are a product of intermediate-mass stars which undergo binary evolution, they do not contribute substantially to chemical enrichment for $\sim 10^8$ years after the commencement of star formation \citep{Matteucci1986-SupernovaEnrichment}.
% The timescale for Fe enrichment in the Galaxy, a key component of chemical evolution models, is directly linked to the progenitor lifetimes of SNe Ia. 
Therefore, the relative abundances of the $\alpha$-elements and Fe as a function of stellar age trace the balance of SN rates over time.

% [What is the delay time distribution]
The {\it delay time distribution} (DTD) refers to the rate of SN Ia events per unit mass of star formation as a function of stellar population age
% The rate of SNe Ia which explode as a function of time after the formation of a hypothetical unit-mass single stellar population is known as the delay-time distribution 
\citep[for a review, see Section 3.5 of][]{Maoz2014-Review}.
When the DTD is convolved with the Galactic star formation rate (SFR), it yields the overall SN Ia rate. 
The quantitative details of the relationship between \aFe and [Fe/H] are set by the DTD, and as such it is a key parameter in GCE models.
However, the DTD remains poorly constrained because it reflects the detailed evolution of the SN Ia progenitor systems, so different models for the progenitors of SNe Ia will naturally predict different forms for the DTD.

% [Introduction to SNe Ia]
The explosion mechanism(s) of SNe Ia are not fully understood \citep[for reviews, see][]{Maoz2014-Review,Livio2018-ProgenitorReview,Ruiter2020-ProgenitorReview}. Two general production channels have been proposed. In the single-degenerate (SD) case, the WD accretes mass from a close non-degenerate companion until it surpasses $\sim1.4$ M$_\odot$ and explodes \citep{Whelan1973-SDModel,Nomoto1982-SDModel,Yoon2003-SDModel}. In the double-degenerate (DD) case, two WDs merge after a gravitational-wave inspiral \citep{Iben1984-IaBinary,Webbink1984-DDModel,Pakmor2012-WDMerger} or head-on collision \citep{Benz1989-CollisionalDD,Thompson2011-CollisionalDD}. Searches for signs of interaction between the SN ejecta and a non-degenerate companion \citep[e.g.,][]{Panagia2006-RadioEmission,Chomiuk2016-RadioEmission,Fausnaugh2019-EarlyIaLightCurves,Tucker2020-SNeIaSpectra,Dubay2022-SNeIaCSM} or for a surviving companion \citep[e.g.,][]{Schaefer2012-ExCompanionSNR,Do2021-Progenitor1972E,Tucker2023-SN2011fe} have placed tight constraints on the SD channel, heavily disfavoring it as the main pathway for producing ``normal'' SNe Ia. The DD channel, now the preferred model, faces issues of its own with matching observed SN Ia rates because not all WD mergers may lead to a thermonuclear explosion \citep[e.g.,][]{NomotoIben1985-DDMergers,SaioNomoto1998-DDMergers,Shen2012-DDMergers}, and the progenitor systems are difficult to detect even within our own Galaxy \citep{RebassaMansergas2019-WhereAreDDProgenitors}.

% [Theoretical DTDs]
% Uncertainties in the progenitor channels of SNe Ia translate into uncertainties in the shape of the DTD. 
% By formulating analytical DTDs for several progenitor models, \citet{Greggio2005-AnalyticalRates} found that the SD channel tends to produce a steeper DTD function than the DD channel, and the median delay time (by which 50\% of SNe Ia have exploded) can vary by a factor of 10 depending on the chosen progenitor scenario and its particular parameters.
Theoretical models have yet to converge on a single prediction for the DTD.
For the DD channel, assumptions about the distribution of post-common envelope separations and the rate of gravitational wave inspiral suggest a broad $\sim t^{-1}$ DTD at long delay times ($\gtrsim 1$ Gyr), but at short delays ($\lesssim 1$ Gyr) the rate is limited by the need to produce two WDs \citep[see Section 3.5 from][]{Maoz2014-Review}.
Higher-order progenitor systems could also produce a $t^{-1}$ DTD \citep{Fang2018-QuadrupleSystems,Rajamuthukumar2023-TripleEvolution}.
The DTD which would result from the SD channel depends greatly on the assumptions of binary population synthesis, but in general is expected to cover a narrower range of delay times and may feature a steep exponential cutoff at the long end \citep[e.g.,][]{Greggio2005-AnalyticalRates}.
% More recently, \citet{Rajamuthukumar2023-TripleEvolution} computed a DTD from the evolution of triple system progenitors. [Expand, connect binary evolution to delay times, look for other theory papers]

% [Observational DTDs]
Surveys of SNe Ia 
% have provided the primary observational constraints on the DTD. This can be done 
can constrain the DTD
by comparing the observed rate of SNe Ia to their host galaxy parameters \citep[e.g.,][]{Mannucci2005-SNRate,Heringer2019-FieldGalaxyDTD} or inferred star formation histories \citep[SFHs; e.g.,][]{Maoz2012-SloanIIDTD}, measuring SN Ia rates in galaxy clusters \citep[e.g.,][]{Maoz2010-ClusterDTD}, or comparing the volumetric SN Ia rate to the cosmic SFH as a whole \citep[e.g.,][]{Graur2014-VolumetricSNIaRates,Strolger2020-ExponentialDTD}. 
Early studies, which had limited sample sizes, produced unimodal \citep{Strolger2004-SNIaProgenitors} or bimodal \citep{Mannucci2006-TwoPopulations} DTDs where the majority of SNe Ia explode within a relatively narrow range of delay times.
More recent studies have recovered broader DTD functions, with many converging on a declining power-law of $\sim t^{-1}$ \citep[e.g.,][]{Maoz2017-CosmicDTD,Castrillo2021-DTD,Wiseman2021-DESRates}, though there is some evidence for a steeper slope in galaxy clusters \citep{Maoz2017-CosmicDTD,FriedmannMaoz2018-ClusterDTD}.
It is especially difficult to constrain the DTD for short delay times \citep{MaozMannucci2012-SNeIaReview} because of the need for SN Ia rates at long lookback times and uncertainties in the age estimates of stellar populations.

@@

The uncertainties in the SN Ia DTD propagate into GCE models. In principle, the observed chemical abundance patterns should therefore contain information about the SN Ia DTD, and by extension their progenitors.
A few studies have compared different DTDs in one-zone chemical evolution models, but their comparisons to abundance data are limited to the local solar neighborhood \citep[e.g.,][]{Andrews2017-ChemicalEvolution,Palicio2023-AnalyticDTD}. 
[Talk about \citep{Matteucci2009-DTDModels} multi-zone models].
\citet{Poulhazan2018-PrecisionPollution} compared six DTDs in a cosmological smoothed-particle hydrodynamics simulation. They found that models with a prompt SN Ia component produced the highest abundance ratios and the narrowest \aFe distributions at late times, although their simulation was not tuned to reproduce the parameters of the Milky Way, and they did not compare to observed chemical abundances. 
The current era of large spectroscopic surveys such as the Apache Point Observatory Galactic Evolution Experiment \citep[APOGEE;][]{Majewski2017-APOGEE} and the forthcoming Milky Way Mapper \citep{Kollmeier2017-SDSS-V} has made abundances across the Milky Way disk available for comparison to more sophisticated GCE models.

% Notes on GCE studies
% \citet{Andrews2017-ChemicalEvolution} predict abundance tracks and distributions using one-zone models for four different DTDs; they find that most are consistent with their abundance data.
% \citet{Palicio2023-AnalyticDTD} derive analytic solutions to one-zone chemical evolution equations for seven DTDs, but their comparisons to data are limited to the solar neighborhood. 
% \citet{Poulhazan2018-PrecisionPollution} investigate six different DTDs in a smoothed-particle hydrodynamics simulation, but they only present global results for the entire galaxy, they don't compare abundances throughout the disk, and they don't compare to observations.

This paper presents the first comprehensive look at the DTD in a multi-zone GCE model that can qualitatively reproduce the observed abundance structure of the Milky Way disk. A multi-zone approach allows for a radially-dependent parameterization of the SFH, outflows, stellar migration, and abundance gradient which can better match observations across the Galactic disk. We evaluate a selection of DTDs from the literature with multiple SFHs and a prescription for radial stellar migration in the Versatile Integrator for Chemical Evolution \citep[\vice;][]{JohnsonWeinberg2020-Starbursts}. In Section \ref{sec:methods}, we present our models for the DTD and SFH. In Section \ref{sec:onezone-results}, we detail our one-zone chemical evolution models and present results. In Section \ref{sec:multizone-results}, we present the results of our multi-zone models and compare to observations. In Section \ref{sec:discussion}, we discuss the implications for the DTD and future surveys. In Section \ref{sec:conclusions}, we summarize our conclusions.

@@

We use \vice to run chemical evolution models which closely follow those of \citet{JohnsonWeinberg2020-Starbursts} and \citet[][hereafter \citetalias{Johnson2021-Migration}]{Johnson2021-Migration}. We refer the interested reader to the former for details about the \vice package and to the latter for details about the model Milky Way disk, including the star formation law, radial density gradient, and outflows. Similar to \citetalias{Johnson2021-Migration}, we adopt a prescription for radial migration based on the {\tt h277} hydrodynamical simulation \citep{Christensen2012-h277}. Appendix \ref{app:migration} describes our method for determining the migration distance $\Delta R_{\rm gal}$ for each model stellar population, which produces smoother distributions in chemical abundance space than the simulation-based approach.

@@

For simplicity and easier comparison to the results of \citetalias{Johnson2021-Migration}, we focus our analysis on O and Fe, representing the $\alpha$ and Fe-peak elements, respectively. Both elements are produced by CCSNe. \vice adopts the instantaneous recycling approximation for CCSNe, so the equation which governs CCSN enrichment as a function of star formation for some element $x$ is simply
\begin{equation}
    \dot M_x^{\rm CC} = y_x^{\rm CC} \dot M_*
    \label{eq:ccsn-enrichment}
\end{equation}
where $y_x^{\rm CC}$ is the CCSN yield of element $x$ per unit mass of star formation, and $\dot M_*$ is the SFR. 
Following \citetalias{Johnson2021-Migration}, who in turn adopt their CCSN yields from \citet{ChieffiLimongi2004-CCSNYields} and \citet{LimongiChieffi2006-CCSNYields}, we adopt $y_{\rm O}^{\rm CC}=0.015$ and $y_{\rm Fe}^{\rm CC}=0.0012$. The primary effect of these yields is to set the low-[Fe/H] ``plateau'' in [O/Fe] which represents pure CCSN enrichment. The chosen yields for this paper produce a plateau at ${\rm [O/Fe]}=0.45$; see \citet{Weinberg2023-CCSNYield} for more discussion on the effect of the CCSN yields on chemical evolution.

Following the formalism of \citet{Weinberg2017-ChemicalEquilibrium}, the rate of Fe contribution to the ISM from SNe Ia is 
\begin{equation}
    \dot M_{\rm Fe}^{\rm Ia} = y_{\rm Fe}^{\rm Ia} \langle \dot M_*\rangle_{\rm Ia} \\
\end{equation}
where $\langle \dot M_*\rangle_{\rm Ia}$ is the time-averaged SFR weighted by the DTD and $y_{\rm Fe}^{\rm Ia}$ is the Fe yield of SNe Ia. \citet{Weinberg2017-ChemicalEquilibrium} show in their Appendix A that
\begin{equation}
    \langle \dot M_*\rangle_{\rm Ia} \equiv \frac{\int_0^t \dot M_*(t')R_{\rm Ia}(t-t')dt'}{\int_{t_D}^{t_{\rm max}} R_{\rm Ia}(t')dt'}
    \label{eq:weighted-sfr}
\end{equation}
where %$\dot M_*(t)$ is the SFR at time $t$, 
%$R_{\rm Ia}(t) dt$ is the number of SNe Ia which explode in the time interval $(t, t+dt)$ per unit mass of stars formed at $t=0$, 
$R_{\rm Ia}$ is the DTD in units of ${\rm M}_\odot^{-1}\,{\rm yr}^{-1}$,
$t_D$ is the minimum SN Ia delay time, and $t_{\rm max}$ is the maximum simulation time. The denominator of Equation \ref{eq:weighted-sfr} is therefore the total number of SNe Ia per M$_{\odot}$ of stars formed $N_{\rm Ia}/M_*$.

The yield $y_{\rm Fe}^{\rm Ia}$ measures the mass of Fe produced by SNe Ia over the full duration of the DTD, which can be expressed as:
\begin{equation}
    y_{\rm Fe}^{\rm Ia} = m_{\rm Fe}^{\rm Ia} \int_{t_D}^{t_{\rm max}} R_{\rm Ia}(t')dt' = m_{\rm Fe}^{\rm Ia} \frac{N_{\rm Ia}}{M_*},
    \label{eq:dtd-integral}
\end{equation}
where $m_{\rm Fe}^{\rm Ia}$ is the average mass of Fe produced by a single SN Ia, and $N_{\rm Ia}/M_*=2.2\pm1\times10^{-3}\,{\rm M}_\odot^{-1}$ is the average number of SNe Ia per mass of stars formed \citep{MaozMannucci2012-SNeIaReview}. 
Adjusting the value of $y_{\rm Fe}^{\rm Ia}$ primarily affects the end point of chemical evolution tracks in [O/Fe]--[Fe/H] space. 
Following \citetalias{Johnson2021-Migration}, we adopt $y_{\rm Fe}^{\rm Ia}=0.00214$. This yield is originally adapted from the W70 model of \citet{Iwamoto1999-SNIaYields}, but it is increased slightly so that the inside-out SFH produces stars with ${\rm [O/Fe]}\approx 0.0$ by the end of the model.
\citet{Palla2021-SNIaYield} studied the effect of different SN Ia yields on GCE models in detail.

\subsection{Delay Time Distributions}
\label{sec:dtd-models}

We explore five different functional forms for the DTD: a single power-law, a broken power-law with an initially flat plateau, a simple exponential, a two-population model, and a model based on triple-system dynamics. We also investigate one or two useful variations of the input parameters for each functional form. Figure \ref{fig:dtds} presents a selection of these DTDs, and Table \ref{tab:dtds} summarizes all of our DTDs and their parameters. We use simple forms rather than simulated physical or analytical models of SNe Ia for the sake of decreased computational time and easier interpretation of the model predictions. Physically-motivated models of the DTD must contend with many unknown or poorly-constrained parameters, so our simplified forms have the advantage of reducing the number of free parameters. In Appendix \ref{app:analytical-dtds}, we show that a few of our simple forms adequately approximate the more complete analytic models of \citet{Greggio2005-AnalyticalRates}.
